Modelling the creation of ecological traps by environmentally persistent pathogens

### 2015-06-01 Manuscript/manuscript.tex
Reorganization of discussion

This work demonstrates that the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting patterns of host occupancy.  
These impacts are driven largely by differences in the colonization and extinction rates of high and low quality habitat, which in turn drive how low and high quality patches interact with and modulate pathogen transmission, either augmenting it in the case of high quality patches, or dampening it in the case of low quality habitat.

@@

@@

These findings that low quality habitat inhibits disease spread while high quality habitat enhances it suggest the presence of a trade-off between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
In this light, this work expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).  
Shifting the patch quality distribution towards high quality habitat in our model is  conceptually similar to increasing movement in these models, and as in these models, the net consequences of this trade-off for the host metapopulation depend on the traits of the pathogen.
Broadly, we see three different scenarios: pathogen traits produce very weak pathogen spread and high quality habitat is beneficial; pathogen traits facilitate intermediate/endemic pathogen spread and low quality habitat is beneficial; and pathogen traits produce easy/pandemic pathogen spread and high quality habitat is again beneficial.

Similarly, in the third scenario, when the pathogen spreads easily as a result of low disease-induced mortality and high transmission (upper portions of panels in Fig~\ref{highvlow}), high quality habitat is again a net benefit to the total host population size.  
Under these conditions, habitat quality again has a limited effect on epidemic dynamics, as the pathogen spreads easily enough to overwhelm any metapopulation structure.
Thus, the benefit of high quality habitat derives primarily from its ability to maintain metapopulation stability in the face of widespread disease-induced mortality.

However, in the second scenario, where the pathogen's traits facilitate intermediate spread (i.e. at an endemic level), the distribution of habitat quality exerts a large influence on epidemic dynamics.
In these cases, the presence of high quality habitat functions similarly to increasing movement in the model developed by Hess (\cite{Hess1996}): increasing pathogen spread and as a consequence reducing overall host population size (Fig~\ref{highvlow}, \ref{popsizes}).  
Here, the consequences of increased pathogen spread outweigh the positive consequences of increased metapopulation connectivity and stability.  
High quality habitat distributions require lower pathogen longevities for pathogen spread to begin to take off, relative to low quality habitat distributions.  Thus, for intermediate longevities, high quality distributions experience considerably greater pathogen spread and lower occupancy than low quality distributions.

@@

>>>>

\subsection*{Introduction and outline of 3 regions}

This work demonstrates that for many pathogens, the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting patterns of host occupancy.  
These impacts are driven largely by the interaction between pathogen traits (i.e. disease-induced mortality, transmission rates) and the colonization and extinction rates of high and low quality habitat.

The general findings that low quality habitat can inhibit disease spread while high quality habitat can enhance it suggest the presence of a trade-off between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
In this light, this work expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).  
Shifting the patch quality distribution towards high quality habitat in our model is  conceptually similar to increasing movement in these models, and as in these models, the net consequences of this trade-off for the host metapopulation depend on the traits of the pathogen.
Broadly, we see three different scenarios: pathogen traits produce very weak pathogen spread and high quality habitat is a net benefit; pathogen traits facilitate intermediate/endemic pathogen spread and low quality habitat is a net benefit; and pathogen traits produce easy/pandemic pathogen spread and high quality habitat is again a net benefit.

\subsection*{Pathogens for which high quality beneficial}

Similarly, in the third scenario, when the pathogen spreads easily as a result of low disease-induced mortality and high transmission (upper portions of panels in Fig~\ref{highvlow}), high quality habitat is again a net benefit to the total host population size.  
Under these conditions, habitat quality again has a limited effect on epidemic dynamics, as the pathogen spreads easily enough to overwhelm any metapopulation structure.
Here then, the benefit of high quality habitat derives primarily from its ability to maintain metapopulation stability in the face of widespread disease-induced mortality.

\subsection*{Pathogens for which high quality detrimental}

However, in the second scenario, where the pathogen's traits facilitate intermediate spread (i.e. at an endemic level), the distribution of habitat quality exerts a large influence on epidemic dynamics. 
Here, the consequences of increased pathogen spread outweigh the positive consequences of increased metapopulation connectivity and stability.

%In these cases, the presence of high quality habitat functions similarly to increasing movement in the model developed by Hess (\cite{Hess1996}): increasing pathogen spread and as a consequence reducing overall host population size (Fig~\ref{highvlow}, \ref{popsizes}). 
%High quality habitat distributions require lower pathogen longevities for pathogen spread to begin to take off, relative to low quality habitat distributions.  Thus, for intermediate longevities, high quality distributions experience considerably greater pathogen spread and lower occupancy than low quality distributions.

\subsubsection*{Role of high quality}

@@

\subsubsection*{Role of low quality}

@@

\subsubsection*{Overall consequences}

@@

\subsubsection*{Conclusions}
